Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-17 article.tex
improve wording, grammar on first four pages of article

In addition to recognizing a stream of tokens and building a parse tree, a
real-world parser must recognize faulty input as such, and accurately report the
location and cause of the failure.  But how -- what generates a parse error,
what information is included in the error, how do error-reporting parsers
compose?

Monad transformers provide a clean and simple answer to these questions.
We'll use them to build a parser combinator library, then we'll use the library
build a simple parser, and then we'll add in error reporting.
\end{introduction}

\section{Parser Combinators}
Parser combinators are an excellent approach for building parsers that are easy
to maintain and expressive.  In addition, they benefit from host-language
integration, allowing them to snarf features such as type systems.

There are such many excellent papers for reading about parser combinators, such
as this \cite{wadler}, this \cite{hutton}, and this \cite{leijen}.  In fact, the
inspiration and terminology for much of this article is drawn directly from
those references.

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the real stuff.  Without any 
further ado, here's the code:
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

@@

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we define the Parser type, which says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and succeed otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
For choosing between alternatives, the Alternative type class will do, so we create
an instance that is defined partly in terms of the Alternative instance of the
underlying result type, \verb+Maybe+.
Last are a couple of convenience combinators for creating specific token parsers.

@@

We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second.  Note that the token position is handled such that the
second parser sees the token list output by the first parser, and that both parsers
must succeed in order for the entire parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

\section{Monad transformers}
Monad transformers \cite{liang} allow monads to be modularly combined to form new monads
which combine the semantics of their constituents.  That's exactly what we need
to do to build parsers.
We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
since they're easy to obtain, easy to use, and quite effective in practice, but
occasionally we'll have to make an extension to avoid certain type class
instances.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe, so the new
parser type is:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}

Now we can build the \verb+item+ parser using the combinators from MonadState, 
since its token lists are the 'state' that it modifies.
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

>>>>

In addition to recognizing a stream of tokens and building a parse tree, a
real-world parser must recognize faulty input, and accurately report the
location and cause of the problem.  But how -- what generates a parse error,
what information is included in the error, how do error-reporting parsers
compose?

Monad transformers provide a clean and simple answer to these questions.
We'll use them to build a parser combinator library, then we'll use the library
to build a simple parser, and finally we'll add in error reporting.
\end{introduction}

\section{Parser Combinators}
Parser combinators are an excellent approach for building expressive parsers 
that are easy to maintain.  In addition, they benefit from host-language
integration, allowing them to snarf features such as type systems and test 
frameworks.

There are many excellent papers on parser combinators, such
as this \cite{wadler} classic Wadler paper on non-deterministic parser combinators, 
Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
and Leijen's Parsec paper \cite{leijen}.  
The inspiration and and much of the names used in this article are drawn 
from those references.

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the hard stuff.  Without any 
further ado, here's the code:
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

@@

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we define the Parser type, which says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and fail otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
For choosing between alternatives, the Alternative type class will do, so we create
an instance that is defined partly in terms of the Alternative instance of the
underlying result type, \verb+Maybe+.
Last are a couple of convenience combinators for creating specific token parsers.

@@

We can run parsers in sequence using the Applicative combinators.  First one parser
is run, then the second, and the token position is threaded between the two.
Note that both parsers must succeed in order for the combined parser to succeed:
\begin{verbatim}
*Main> getParser (item *> item) "abcde"
Just ("cde",'b')

@@

\section{Monad transformers}
Monad transformers \cite{liang} allow monads to be modularly combined to form 
new, more complicated monads; the combined monads support all the operations of
their constituents.  Doing just that -- combining multiple monads into a stack --
we can build semantically rich yet modular parsers.
We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
since they're easy to obtain, easy to use, and quite effective in practice, but
occasionally we'll have to make a slight modification or extensions.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.  We can 
now define the new parser type as:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}

We can build the \verb+item+ parser using the combinators from MonadState --
token lists are the 'state' that \verb+item+ must inspect and modify:
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}
